ANTS: Advanced Open-Source Tools for Normalization And Neuroanatomy

### 2009-09-30 Documentation/Tex/antstheory.tex
update refs
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@58 fd2da0a4-d968-48df-9cb5-d8c496287cbf
ok - minor improvements
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@160 fd2da0a4-d968-48df-9cb5-d8c496287cbf

\begin{abstract}
Computational anatomy (CA) seeks to quantify natural variation in
biological shape and function with roots that reach back to the
seminal works of Charles Darwin and D'arcy Thompson.  CA is currently
applied to study health, disease and epidemiology and uses deformable
mappings between images as a basic technique.  However, there is a
lack of standards and reproducibility in the field that is due, in
part, to the use of proprietary software and private data.  To
facilitate reproducibility in CA measurements and advancement of
imaging sciences, NIH has recently committed significant support to
open-source data and software resources.  Here, we report a recent
product of this commitment: Advanced Normalization Tools (ANTS), an
ITK-based toolkit for CA and related areas.  The ANTS open-source
library consists of a well-evaluated suite of state-of-the-art
normalization and template-building tools for quantitative
morphometric analysis.  We highlight the prominent features of ANTS
and demonstrate its utility by performing a detailed analysis on
openly-available anatomically labeled brain data from the non-rigid
image registration evaluation project (NIREP).  The results from this
analysis evidences the high level of accuracy achievable with ANTS
using intensity-based registration alone.  In addition, we show the
significant performance gains may be achieved by coupling
intensity-based image metrics and point-set-based metrics from
specific, sensibly selected cortical structures.
%Computational anatomy is the quantitative study of biological shape and its natural variations which traces its roots to the seminal works of Charles Darwin and D'arcy Thompson.  Integral to this research field is the calculation of deformable mappings between images that serve as invaluable measurement tools.  Technological evolution has spawned much innovation in the computational development of such deformable mapping algorithms.  However, not only is algorithmic development essential for increasing scientific knowledge but the availability of these technologies is equally crucial for facilitating the progress of the scientific community at large.  This is the motivation behind such projects as the Insight Toolkit (ITK) of the National Institutes of Health which serves as a repository for many popular image segmentation and registration routines.  In addition, the ITK library provides a well-structured framework for several stand-alone projects such as Elastix, ITK-SNAP, ManagedITK, the Medical Imaging Toolkit (MITK), WrapITK, and many others.  Our own  contribution to the field of computational anatomy, also built on an ITK foundation, is called Advanced Normalization Tools (ANTS).  The ANTS open-source library consists of an entire suite of state-of-the-art normalization and template-building tools for quantitative morphometric analysis. After highlighting the prominent features of ANTS, we demonstrate its utility by performing a detailed analysis on anatomically labeled brain data from the non-rigid image registration evaluation project (NIREP).  The results from this analysis evidences the high level of accuracy achievable with ANTS using intensity-based registration alone.  In addition, we show the performance gains achieved by coupling intensity-based image metrics and point-set-based metrics from specific, sensibly selected cortical structures.

%In a recent evaluation of 14 non-linear deformation algorithms \cite{Klein2009}, the SyN algorithm of ANTS was consistently one of the top performers.

\end{abstract}

\IEEEkeywords{
ANTS, image registration, nueroanatomy, neuroimaging, normalization, open-source, point-set registration
}

\IEEEpeerreviewmaketitle

@@

The rapid advancement of biological and medical imaging technologies has caused a proliferation in the development of quantitative tools for computational anatomy.  The principal tools of this emerging field are meaningful deformable mappings between images whether they be driven by similarity metrics which are intensity-based,  point-set based, or both.  Several categories of mappings have been proposed in the literature.  Of particular recent interest are diffeomorphic transformations which, by definition, preserve topology.  Topology preservation is fundamental to making comparisons between objects in the natural world as such transformations permit comparisons to be made across time points in an individual's disease process or to study development patterns across a large population.

@@

We first provide an overview of the different components included in ANTS, such as the available transformation models and similarity metrics offered.  This is followed by an extensive experimental analysis that builds upon the results from the recent image normalization evaluation of \cite{Klein2009} which was limited to a single configuration of ANTS.  This will provide  a useful benchmark for evaluating other possible ANTS configurations.

@@

A variety of transformation models have been proposed in the literature with varying degrees of freedom (illustrated in Figure \ref{fig:hierarchy}).  For deformable transformations, one approach is to optimize within the space of non-topology preserving, yet physically motivated transformations---an approach pioneered by Bajcsy \cite{Bajcsy1989}.  Elastic and related models, such as HAMMER \cite{Shen2002}, statistical parametric mapping (SPM) \cite{Ashburner2000}, free-form deformations (FFD) \cite{Rueckert1999}, and
Thirion's demons \cite{Thirion1998} operate in the space of vector fields, which does not preserve topology.  In other words, barring ad hoc constraints to prevent otherwise, these algorithms allow the topology to change in an uncontrolled way which makes the deformable mappings difficult to interpret in functional or anatomical studies.

These topology concerns have led to current advocation within the computational anatomy research community of diffeomorphic mappings for transformation modeling.   Optimizing directly within this space has shown remarkable success in various computational anatomy studies involving longitudinal \cite{Avants2007b,Fox2001}, functional \cite{Miller2005}, and population data \cite{Avants2007b}.  We include three such diffeomorphic algorithms in the ANTS toolbox based on previously reported research.

@@

\subsubsection{Vector Field Operators for Regularization} Antecedent to discussion of available deformable transformation models, we point out that  deformable normalization strategies typically invoke a deformation regularization step which smooths the displacement field, $\displace$, or velocity field, $\mathbf{v}$,  or both by a linear operator such as the Laplacian or Navier-Stokes operator. One may write this regularization as a variational minimization in terms of its linear operator or in
terms of a kernel function operating of the field itself, e.g., $
\displace_{smooth}= K \star \displace_{not \,\, smooth}$, where $K \star$ denotes 
convolution with the Green's kernel, $K$, for the linear operator, $L$.

@@

\subsubsection{Diffeomorphic Transformations}

In contrast to many transformation models which reside in the domain of vector spaces, a diffeomorphism is a differentiable mapping with a differentiable inverse \cite{Ebin1970,Mumford1998}.  Modeling transformations with diffeomorphisms ensures certain desirable topological properties that cannot be guaranteed with other methods.

ANTS assumes the diffeomorphism, $\phi$, is defined on the image domain, $\Omega$, an maintains an affine transform at the boundary such that $\phi(\partial \Omega) = A(\mathbf{Id})$ where $A(\mathbf{Id})$ is an affine mapping applied to the identity transformation.  $\phi$, over time, parameterizes a family of diffeomorphisms, $\phi(\mathbf{x}, t) : \Omega \times t \rightarrow  \Omega$, which can be generated by integrating a time-dependent, smooth velocity field, $\boldsymbol{v} : \Omega \times t \rightarrow \mathbb{R}^d$,  through the ordinary differential equation (o.d.e.)
\begin{align} \label{eq:ode}
\frac{d \phi(\mathbf{x}, t)}{dt} = \boldsymbol{v}(\phi(\mathbf{x}, t), t), \,\,\, \phi(\mathbf{x}, 0) = \mathbf{x}.
\end{align}
The existence and uniqueness theorem for o.d.e.'s implies that integrating Equation (\ref{eq:ode}) generates a diffeomorphism.  The deformation field yielded by $\phi$ is $\mathbf{u}(\mathbf{x}) = \phi(\mathbf{x},1) - \mathbf{x}$.

Dupuis et al. \cite{Dupuis1998} motivated the usage of diffeomorphisms for CA  by showing that the variational form
\begin{align}\label{eq:D}
  D(\mathcal{I}, \mathcal{J}) =  \int_0^1 ||L\boldsymbol{v}||^2 dt , \,\, \mathcal{I}(\phi(\mathbf{x}, 1)) = \mathcal{J}(\mathbf{z})
\end{align}
represents a true mathematical metric between anatomical instances $\mathcal{I}$ and $\mathcal{J}$ given an appropriate norm, $L$, on the velocity field, $\boldsymbol{v}$.  An optimal solution, $\boldsymbol{v}^*$, minimizes the metric $D(\mathcal{I}, \mathcal{J})$ with respect to $L$.   Dupuis \cite{Dupuis1998} also showed that such a solution is guaranteed to be well-posed.  Intuitively, Equation (\ref{eq:D}) provides a sense of distance between two anatomical shapes.  It also illustrates that the optimal diffeomorphic solution is analogous to finding the geodesic curve between two points in a curved space.%
\footnote{
It is important to note the similarity between the definition of curve length, $\int ||\mathcal{C}'(t)||dt$, for the parametric curve $\mathcal{C}(t)$ and Equation (\ref{eq:D}).  In this sense, the solution for Equation (\ref{eq:D}) is the geodesic diffeomorphism, where $\boldsymbol{v}$ is the tangent vector of the diffeomorphism, such that the shape distance, $D$, between $\mathcal{I}$ and $\mathcal{J}$ is minimized.
}

@@

Exploiting the fact that the diffeomorphism, $\phi$, can be decomposed into two components 
$\phi_1$ and $\phi_2$, \cite{Avants2007b} construct a {\em symmetric} alternative to Equation (\ref{eq:diff}).  This leads to the symmetric variant of Equation (\ref{eq:diff})
\begin{align}\label{eq:diffs}
  \{\boldsymbol{v}_1^*,\boldsymbol{v}_2^*\}  = &
     \underset{\boldsymbol{v_1},\boldsymbol{v_2}}{\operatorname{argmin}}  
      \Bigg\{ \nonumber \\
      & \int_0^{0.5} \left( ||L\boldsymbol{v}_1(x,t)||^2 + ||L\boldsymbol{v}_2(x,t)||^2   \right)dt  \nonumber \\
     &+ \lambda\int_{\Omega}\Pi_{\sim}\left( \mathcal{I} \circ \phi_1(\mathbf{x},0.5), \mathcal{J}\circ \phi_2(\mathbf{x},0.5)\right)d\Omega\Bigg\}.
\end{align}
The corresponding symmetric Euler-Lagrange equations are similar to \cite{Miller2002}.  Finding $\boldsymbol{v_1}^*$ minimizes the variational energy from $t = 0$ whereas $\boldsymbol{v_2}^*$ minimizes from $t = 1$.  Thus, gradient-based iterative convergence deforms $\mathcal{I}$ and $\mathcal{J}$ along the geodesic diffeomorphism, $\phi$, to a fixed  point midway (intuited by the notion of shape distance) between $\mathcal{I}$ and $\mathcal{J}$ thus motivating the denotation of the solution strategy as Symmetric Normalization (SyN).

@@

\paragraph{Geodesic SyN}
Using a gradient-based optimization strategy, minimization implied by Equation (\ref{eq:diffs}), which we denote as $\Pi$, first requires a specified discretization of $t \in [0,1]$.  Calculation of the gradient of $\Pi$ with respect to the diffeomorphisms $\phi_1$ and $\phi_2$ is performed at each of the $t$ values which is given by
\begin{align}
  \nabla \Pi(\boldsymbol{x},t) = \frac{\partial}{\partial \phi_i} \Pi_{\sim}( \mathcal{I}(\phi_{1}^{-1}(\boldsymbol{x},t)), \mathcal{J}(\phi_{2}^{-1}(\boldsymbol{x},1-t)) 
\end{align}
for $i \in \{1,2\}$.   The corresponding velocity fields at each $t$ are then updated from the previous iteration according to 
\begin{align}
  \boldsymbol{v}(\mathbf{x},t) = \boldsymbol{v}(\mathbf{x},t) + K \star \nabla \Pi(\mathbf{x},t).
\end{align}
We then integrate Equation (\ref{eq:ode}) using Runge-Kutta methods to yield $\phi_i(\mathbf{x},t)$ for each $t \in [0,1]$ and $i \in \{1,2\}$.  We cycle through these steps until convergence or iterative exhaustion.

@@

Greedy optimization of Equation (\ref{eq:diffs}) requires calculation of the gradient of the gradient taken at the mid-point of the full diffeomorphism, i.e. at $t = 0.5$
\begin{align}
\nabla \Pi =\frac{\partial}{\partial \phi_i} \Pi_{\sim}( \mathcal{I}(\phiinv_1(\x,0.5)) ,  \mathcal{J}(\phiinv_2(\x,0.5)) )
\end{align}
for $i \in \{1,2\}$.  $\phi_1(\mathbf{x},0.5)$ and $\phi_2(\mathbf{x},0.5)$ are then updated from the previous iteration according to 
\begin{align}
  \phi_i(\mathbf{x},0.5) = \phi_i(\mathbf{x},0.5)+\delta  (K \star \nabla \Pi ( \phi_i(\x,0.5) )).
\end{align}
where $\delta$ is a user-specifed step parameter.  The gradient is then mapped back to the origin of each diffeomorphism to update the full mapping by enforcing $\phiinv(\phi(\x,1))=\x$.

@@

Based on our experience with standard command line argument parsing packages (e.g. \verb#getopt#), we developed our own set of classes for an intuitive command line interface.  A summary of command line arguments are given in Table \ref{table:args}.  
These ANTS argument parsing classes provide an intuitive compromise between parsers where every variable requires a unique flag and strict ordering requirements on the command line.  A typical command line call to ANTS  is  given by
\begin{lstlisting}
>ANTS 3 
  --metric MSQ[fixedImage.nii,movingImage.nii,0.75] 
  --metric PSE[fixedPoints.vtk,movingPoints.vtk,0.25] 
  --transformation Exp[0.75] 
  --regularization Gauss[3.0,0.5]
  --iterations 50x20x10x5 
  --output-naming results.nii
\end{lstlisting}
The correspondence between the ANTS command line specification and the image normalization formulation with labeled point-sets $X$ and $Y$ illustrates the motivation for our command line interface,
\begin{align}
  \Pi(&\mathcal{I},\mathcal{J},X,Y,\phi) \nonumber \\ 
     =& \underbrace{\int_0^1 ||L\boldsymbol{v}||dt}_{\texttt{-t Exp[}\cdot \texttt{]},\texttt{ -r Gauss[}\cdot\texttt{,}\cdot\texttt{]}}  + \underbrace{\lambda_1\int_{\Omega} || \mathcal{I} \circ \phi(\mathbf{x},t) - \mathcal{J}  || d\Omega}_{\texttt{-t Exp[}\cdot \texttt{]},\texttt{ -m MSQ[}\mathcal{J}\texttt{,}\mathcal{I}\texttt{,}\lambda_1\texttt{]}} \nonumber \\
&+\underbrace{\frac{\lambda_2 }{|Y|}\sum_{i=1}^{|Y|} \left\|y_i - \frac{1}{|X|} \sum_{j = 1}^{|X|} G(y_i; x_j, \sigma_X) x_j\right\|^2}_{\texttt{-m PSE[}Y\texttt{,}X\texttt{,}\lambda_2\texttt{]}}.
\end{align}

@@

% use section* for acknowledgement
\section*{Acknowledgment}
ANTS is supported by Grant 1R01EB006266-01 from the National Institute Of Biomedical Imaging and Bioengineering and administered through the UCLA Center for Computational Biology.
% optional entry into table of contents (if used)
%\addcontentsline{toc}{section}{Acknowledgment}
%The authors would like to thank...

% trigger a \newpage just before the given reference
% number - used to balance the columns on the last page
% adjust value as needed - may need to be readjusted if
% the document is modified later
%\IEEEtriggeratref{8}
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% references section
% NOTE: BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/

% can use a bibliography generated by BibTeX as a .bbl file
% standard IEEE bibliography style from:
% http://www.ctan.org/tex-archive/macros/latex/contrib/supported/IEEEtran/bibtex
%\bibliographystyle{IEEEtran.bst}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)

\bibliographystyle{IEEEtran.bst}
\bibliography{./references.bib}

>>>>

\begin{abstract}
Computational anatomy (CA) seeks to quantify natural variation in
biological shape and function with roots that reach back to the
seminal works of Charles Darwin and D'arcy Thompson.  CA is currently
applied to study health, disease and epidemiology and uses deformable
mappings between images as a basic technique.  However, there is a
lack of standards and reproducibility in the field that is due, in
part, to the use of proprietary software and private data that is difficult 
to openly evaluate.  To
facilitate reproducibility in CA measurements and advancement of
imaging sciences, NIH has recently committed significant support to
open-source data and software resources.  Here, we report a recent
product of this commitment: Advanced Normalization Tools (ANTS), an
ITK-based toolkit for CA and related areas.  The ANTS open-source
library consists of a well-evaluated suite of state-of-the-art
normalization and template-building tools for quantitative
morphometric analysis.  We highlight the prominent features of ANTS
and demonstrate its utility by performing a detailed analysis on
openly-available anatomically labeled brain data from the non-rigid
image registration evaluation project (NIREP).  The results from this
analysis evidences the high level of accuracy achievable with ANTS
using intensity-based registration alone.  In addition, we show the
significant performance gains may be achieved by coupling
intensity-based image metrics and point-set-based metrics from
specific, sensibly selected cortical structures.
%Computational anatomy is the quantitative study of biological shape and its natural variations which traces its roots to the seminal works of Charles Darwin and D'arcy Thompson.  Integral to this research field is the calculation of deformable mappings between images that serve as invaluable measurement tools.  Technological evolution has spawned much innovation in the computational development of such deformable mapping algorithms.  However, not only is algorithmic development essential for increasing scientific knowledge but the availability of these technologies is equally crucial for facilitating the progress of the scientific community at large.  This is the motivation behind such projects as the Insight Toolkit (ITK) of the National Institutes of Health which serves as a repository for many popular image segmentation and registration routines.  In addition, the ITK library provides a well-structured framework for several stand-alone projects such as Elastix, ITK-SNAP, ManagedITK, the Medical Imaging Toolkit (MITK), WrapITK, and many others.  Our own  contribution to the field of computational anatomy, also built on an ITK foundation, is called Advanced Normalization Tools (ANTS).  The ANTS open-source library consists of an entire suite of state-of-the-art normalization and template-building tools for quantitative morphometric analysis. After highlighting the prominent features of ANTS, we demonstrate its utility by performing a detailed analysis on anatomically labeled brain data from the non-rigid image registration evaluation project (NIREP).  The results from this analysis evidences the high level of accuracy achievable with ANTS using intensity-based registration alone.  In addition, we show the performance gains achieved by coupling intensity-based image metrics and point-set-based metrics from specific, sensibly selected cortical structures.

%In a recent evaluation of 14 non-linear deformation algorithms \cite{Klein2009}, the SyN algorithm of ANTS was consistently one of the top performers.
{\bf FIXME -- make clear what transformations we are comparing with constant similarity ... }
\end{abstract}

%\IEEEkeywords{
%ANTS, image registration, nueroanatomy, neuroimaging, normalization, open-source, point-set registration
%}

%\IEEEpeerreviewmaketitle
\maketitle

@@

The rapid advancement of biological and medical imaging technologies
has caused a proliferation in the development of quantitative tools
for computational anatomy.  The principal tools of this emerging field
are deformable mappings between images whether they be driven by
similarity metrics which are intensity-based, point-set based, or
both.  Several categories of mappings have been proposed in the
literature.  Of particular recent interest are diffeomorphic
transformations which, by definition, preserve topology.  Topology
preservation is fundamental to making comparisons between objects in
the natural world that are thought to change in such a way that local
neighborhoods are preserved.  Cytoarchitectonic brain mapping studies
also suggest that the layout of cell types throughout the brain is
generally preserved \cite{Schleicher2009}, further motivating diffeomorphic mapping
in the context of the brain.

@@

We first provide an overview of the different components included in
ANTS, such as the available transformation models and similarity
metrics offered.  This is followed by an extensive experimental
analysis that builds upon the results from the recent image
normalization evaluation of \cite{Klein2009} which was limited to a
single configuration of ANTS.  This will provide a useful benchmark
for evaluating other possible ANTS configurations.  Importantly, this
work also suggests---to our knowledge, for the first time---the clear
practical benefit of time-parameterized diffeomorphic mapping over the
greedy and exponential mapping approaches.

@@

A variety of transformation models have been proposed in the literature with varying degrees of freedom (illustrated in Figure \ref{fig:hierarchy}).  For deformable transformations, one approach is to optimize within the space of non-topology preserving, yet physically motivated transformations---an approach pioneered by Bajcsy \cite{Bajcsy1989}.  Elastic and related models, such as HAMMER \cite{Shen2002}, statistical parametric mapping (SPM) \cite{Ashburner2000}, free-form deformations (FFD) \cite{Rueckert1999}, and
Thirion's demons \cite{Thirion1998} operate in the space of vector fields, which does not preserve topology.  In other words, barring ad hoc constraints to prevent otherwise, these algorithms may allow the topology to change in an uncontrolled way which makes the deformable mappings difficult to interpret in functional or anatomical studies.

Thus, in addition to shape-based, biological motivation, diffeomorphic
mapping is motivated by the desire for deformable transformations that
provide well-behaved solutions with mathematical guarantees about
distances and regularity.  Furthemore, the diffeomorphic space has
group structure \cite{Arnold1991}.  Optimizing directly within this
space has shown remarkable success in various computational anatomy
studies involving longitudinal \cite{Avants2007b,Fox2001}, functional
\cite{Miller2005}, and population data \cite{Avants2007b}.  We include
three such diffeomorphic algorithms in the ANTS toolbox based on
previously reported research and a new time-parameterized extension to
the standard symmetric normalization algorithm \cite{Avants2008}.

@@

\subsubsection{Vector Field Operators for Regularization} Antecedent to discussion of available deformable transformation models, we point out that  deformable normalization strategies typically invoke a deformation regularization step which smooths the displacement field, $\displace$, or velocity field, $\mathbf{v}$,  or both by a linear operator such as the Laplacian or Navier-Stokes operator. One may write this regularization as a variational minimization in terms of its linear operator or in
terms of a kernel function operating on the field itself, e.g., $
\displace_{smooth}= K \star \displace_{not \,\, smooth}$, where $K \star$ denotes 
convolution with the Green's kernel, $K$, for the linear operator, $L$.

@@

\subsubsection{Diffeomorphic Transformations}
Diffeomorphisms form a group of differentiable maps with
differentiable inverse \cite{Ebin1970,Mumford1998} that is closed
under composition.  In contrast, the vector space that most deformable
image registration methods use is closed under {\em addition}, an
operation that cannot guarantee topology preservation.  That is, two
topology-preserving vector fields added together are still a vector
field but the result may no longer preserve topology.  Furthermore,
most regularization models for vector spaces use a quadratic penalty,
thus making large deformations difficult to realize.  Modeling
transformations with diffeomorphisms, on the other hand, ensures a
flexible, linear penalty on deformation while guaranteeing topology 
preservation.  Additionally, distance metrics in the space of diffeomorphisms 
allow geodesic properties to be explored \cite{Miller2001,Beg2005}.

ANTS assumes the diffeomorphism, $\phi$, is defined on the image domain, $\Omega$, and maintains an affine transform at the boundary such that $\phi(\partial \Omega) = A(\mathbf{Id})$ where $A(\mathbf{Id})$ is an affine mapping applied to the identity transformation.  $\phi$, over time, parameterizes a family of diffeomorphisms, $\phi(\mathbf{x}, t) : \Omega \times t \rightarrow  \Omega$, which can be generated by integrating a time-dependent, smooth velocity field, $\boldsymbol{v} : \Omega \times t \rightarrow \mathbb{R}^d$,  through the ordinary differential equation (o.d.e.)
\begin{align} \label{eq:ode}
\frac{d \phi(\mathbf{x}, t)}{dt} = \boldsymbol{v}(\phi(\mathbf{x}, t), t), \,\,\, \phi(\mathbf{x}, 0) = \mathbf{x}.
\end{align}
The existence and uniqueness theorem for o.d.e.'s implies that integrating Equation (\ref{eq:ode}) generates a diffeomorphism.  The deformation field yielded by $\phi$ is $\mathbf{u}(\mathbf{x}) = \phi(\mathbf{x},1) - \mathbf{x}$.

Dupuis et al. \cite{Dupuis1998} motivated the usage of diffeomorphisms for CA  by showing that the variational form
\begin{align}\label{eq:D}
  D(\mathcal{I}, \mathcal{J}) =  \int_0^1 ||L\boldsymbol{v}|| dt , \,\, \mathcal{I}(\phi(\mathbf{x}, 1)) = \mathcal{J}(\mathbf{z})
\end{align}
represents a true mathematical metric between anatomical instances $\mathcal{I}$ and $\mathcal{J}$ given an appropriate norm, $L$, on the velocity field, $\boldsymbol{v}$.  An optimal solution, $\boldsymbol{v}^*$, minimizes the metric $D(\mathcal{I}, \mathcal{J})$ with respect to $L$.   Dupuis \cite{Dupuis1998} also showed that such a solution is guaranteed to be well-posed.  Intuitively, Equation (\ref{eq:D}) provides a sense of distance between two anatomical shapes.  It also illustrates that the optimal diffeomorphic solution is analogous to finding the geodesic curve between two points in a curved space.%
\footnote{
It is important to note the similarity between the definition of curve length, $\int ||\mathcal{C}'(t)||dt$, for the parametric curve $\mathcal{C}(t)$ and Equation (\ref{eq:D}).  In this sense, the solution for Equation (\ref{eq:D}) is the geodesic diffeomorphism, where $\boldsymbol{v}$ is the tangent vector of the diffeomorphism, such that the shape distance, $D$, between $\mathcal{I}$ and $\mathcal{J}$ is minimized.
}

@@

Exploiting the fact that the diffeomorphism, $\phi$, can be decomposed into two components 
$\phi_1$ and $\phi_2$, one may construct a {\em symmetric} alternative to Equation (\ref{eq:diff}).  
Now define, in $t \in [0,0.5]$, $\boldsymbol{v}(x,t) = \boldsymbol{v}_1(x,t) $ and 
$\boldsymbol{v}(x,t) = \boldsymbol{v}_2(x,1-t)$ when $t \in [0.5,1]$. 
This leads to the symmetric variant of Equation (\ref{eq:diff}), 
\begin{align}\label{eq:diffs}
  \{ \boldsymbol{v}^{*}_1 , \boldsymbol{v}^{*}_2 \}= &
     \underset{\boldsymbol{v}_{1,2}}{\operatorname{argmin}}  
      \Bigg\{ \nonumber \\
      & \int_0^{0.5} ||L\boldsymbol{v}_1(x,t)||^2 ~dt +  \int_0^{0.5} ||L\boldsymbol{v}_2(x,t)||^2 ~dt  \nonumber \\
     &+ \lambda\int_{\Omega}\Pi_{\sim}\left( \mathcal{I} \circ \phi_1(\mathbf{x},0.5), \mathcal{J}\circ \phi_2(\mathbf{x},0.5)\right)d\Omega\Bigg\}.
\end{align}
Note that the regularization term, here, is equivalent to that in equation~\ref{eq:lddmm}.  The only change is the splitting of the integral into 
two time intervals reflecting the underlying optimized components of the velocity field. 
The corresponding symmetric Euler-Lagrange equations are similar to \cite{Miller2002}.  The difference, here, is that in finding $\boldsymbol{v}^*$, we minimize the variational energy from either end-point towards the mid-point of the transformation, as indicated by the data term.  This strategy ``splits'' the optimization dependence equally between both images.  Thus, gradient-based iterative convergence deforms $\mathcal{I}$ and $\mathcal{J}$ along the geodesic diffeomorphism, $\phi$, to a fixed  point midway (intuited by the notion of shape distance) between $\mathcal{I}$ and $\mathcal{J}$ thus motivating the denotation of the solution strategy as Symmetric Normalization (SyN).

@@

\paragraph{Geodesic SyN}
Using a gradient-based optimization strategy for minimizing equation
(\ref{eq:diffs}) first requires a specified discretization of $t$,
s.t. $t \in [0,1] \approx \{ 0 , 1/k , 2/k , \cdots , 1 \} $ where integer $k > 1$
is the desired number of discretized intervals.  Calculation of the
gradient of $\Pi$ with respect to the diffeomorphisms $\phi_1$ and
$\phi_2$ is performed at each of these $t$ values, $t_k$, 
\begin{align}
\label{eq:grad}
  \nabla \Pi_i(\boldsymbol{x},t_k) = \frac{\partial}{\partial \phi_i} \Pi_{\sim}( \mathcal{I}(\phi_{1}^{-1}(\boldsymbol{x},t_k)), \mathcal{J}(\phi_{2}^{-1}(\boldsymbol{x},1-t_k)) 
\end{align}
for $i \in \{1,2\}$.   The total velocity field is then updated from the previous iteration according to, 
for each $i$, 
\begin{align}
  \boldsymbol{v}(\mathbf{x},t) = \boldsymbol{v}(\mathbf{x},t) + \delta K \star \nabla \Pi_i(\mathbf{x},t),
\end{align}
where $\delta$ is the user-specified gradient descent parameter.  We
then integrate Equation (\ref{eq:ode}) using Runge-Kutta methods to
yield $\phi_i(\mathbf{x},t)$ for each $t \in [0,1]$ and $i \in
\{1,2\}$.  We cycle through these steps until convergence or iterative
exhaustion.

@@

Greedy optimization of Equation (\ref{eq:diffs}) calculates the gradient only at the mid-point of the full diffeomorphism, i.e. at $t = 0.5$ 
in equation~\ref{eq:grad}
% \begin{align}
% \nabla \Pi =\frac{\partial}{\partial \phi_i} \Pi_{\sim}( \mathcal{I}(\phiinv_1(\x,0.5)) ,  \mathcal{J}(\phiinv_2(\x,0.5)) )
% \end{align}
for $i \in \{1,2\}$.  $\phi_1(\mathbf{x},0.5)$ and $\phi_2(\mathbf{x},0.5)$ are then updated from the previous iteration according to 
\begin{align}
  \phi_i(\mathbf{x},0.5) = \phi_i(\mathbf{x},0.5)+ (\delta K \star \nabla \Pi_i(\mathbf{x},0.5) ) \circ  \phi_i(\x,0.5).
\end{align}
That is, the gradient at the mid-point is mapped back to the origin of each diffeomorphism.  
We then update the full mapping by explicitly enforcing $\phiinv(\phi(\x,1))=\x$ in the discrete 
domain, as described in \cite{Avants2008}.

@@

Based on our experience with standard command line argument parsing packages (e.g. \verb#getopt#), we developed our own set of classes for an intuitive command line interface.  A summary of command line arguments are given in Table \ref{table:args}.  
These ANTS argument parsing classes provide an intuitive compromise between parsers where every variable requires a unique flag and strict ordering requirements on the command line.  A typical command line call to ANTS  is  given by
\begin{lstlisting}
>ANTS 3 
  --metric MSQ[fixedImage.nii,movingImage.nii,0.75] 
  --metric PSE[fixedPoints.vtk,movingPoints.vtk,0.25] 
  --transformation Exp[0.75] 
  --regularization Gauss[3.0,0.5]
  --iterations 50x20x10x5 
  --output-naming results.nii
\end{lstlisting}
The correspondence between the ANTS command line specification and the image normalization formulation with labeled point-sets $X$ and $Y$ illustrates the motivation for our command line interface,
\begin{align}
  \Pi(&\mathcal{I},\mathcal{J},X,Y,\phi) \nonumber \\ 
     =& \underbrace{\int_0^1 ||L\boldsymbol{v}||^2 dt}_{\texttt{-t Exp[}\cdot \texttt{]},\texttt{ -r Gauss[}\cdot\texttt{,}\cdot\texttt{]}}  + \underbrace{\lambda_1\int_{\Omega} || \mathcal{I} \circ \phi(\mathbf{x},t) - \mathcal{J}  || d\Omega}_{\texttt{-t Exp[}\cdot \texttt{]},\texttt{ -m MSQ[}\mathcal{J}\texttt{,}\mathcal{I}\texttt{,}\lambda_1\texttt{]}} \nonumber \\
&+\underbrace{\frac{\lambda_2 }{|Y|}\sum_{i=1}^{|Y|} \left\|y_i - \frac{1}{|X|} \sum_{j = 1}^{|X|} G(y_i; x_j, \sigma_X) x_j\right\|^2}_{\texttt{-m PSE[}Y\texttt{,}X\texttt{,}\lambda_2\texttt{]}}.
\end{align}

@@

% use section* for acknowledgement
\section*{Acknowledgment}
ANTS is supported by Grant 1R01EB006266-01 from the National Institute Of Biomedical Imaging and Bioengineering and administered through the UCLA Center for Computational Biology.

\bibliographystyle{IEEEtran.bst}
\bibliography{./references.bib,ants,avantsAll}
